---
title: "Reference corpora, protected benchmarks, and local adapters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reference corpora, protected benchmarks, and local adapters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

LibeRary keeps reference answers outside its live catalogue. This prevents a
benchmark answer from entering retrieval, prompts, demonstrations, or model
training before a protected evaluation.

## Corpus layout and quality tiers

`library_reference_build()` reads the drug-specific v2 systematic-review
supplements. It creates separate article, model, and screening JSON documents,
plus deterministic PMID-level partition indexes. One PMID is always confined
to one partition.

The original appendix row is retained in `raw`. The normalized LibeRary target
is stored under `reference$extraction_target`. Version 0.2.2 also adds structured
cohorts/demographics, dosing, canonical model semantics, inferred or unresolved
ADVAN/TRANS implementations, and reproduction targets. Normalization therefore
remains auditable and never silently changes the historical review.

Reference tiers have different meanings:

- A: directly reported and independently checked;
- B: deterministically derived and independently checked;
- C: appendix reconstruction or assumption requiring source review; and
- D: incomplete or unresolved.

Only A/B records are strict benchmark targets. C/D results remain useful silver
development metrics; additional predictions are called *unverified extras*,
not hallucinations, because an incomplete reference cannot prove absence.

```{r, eval=FALSE}
library(LibeRary)

root <- "validation/liberary/aed-pkpd-reference/0.2.2"
library_reference_build("AED_PKPD", root, version = "0.2.2")
library_reference_validate(root, check_hashes = TRUE,
                           source_dir = "AED_PKPD")
library_reference_list(root)
```

## Target-blind baseline

The runner constructs a source-only work list containing article identity and
PDF location. It never reads `reference$extraction_target` when creating a
prompt. Prediction envelopes explicitly record that safeguard.

```{r, eval=FALSE}
cfg <- ingest_load_config()

library_reference_run(
  root = root,
  source_dir = "AED_PKPD",
  output_dir = "benchmark/aed-current",
  cfg = cfg,
  task = "extraction",
  partition = "test",
  resume = TRUE,
  assess = FALSE
)

score <- library_reference_benchmark(
  root,
  "benchmark/aed-current/predictions",
  task = "extraction",
  partition = "test",
  output_dir = "benchmark/aed-current/report"
)
score$summary
```

Extraction scoring covers normalized scalar fields, model structure, and
THETA/OMEGA/SIGMA/covariance values with declared numeric tolerances. A separate
semantic score covers canonical structure, implementation inference, cohort
counts and demographic summaries, dosing, and any reproduction targets. These
scores remain in the silver tier until their source fields have been reviewed,
so appendix inferences are not treated as strict truth. Triage
scoring reports Brier score, log loss, AUROC, average precision, sensitivity,
and specificity. Wrong-drug and unspecified historic exclusions are omitted
from the general-model triage score because they are query-specific or
ambiguous negatives.

## Side-by-side review

The review application presents three deliberately separate representations:

1. the historical appendix row under `raw`;
2. the normalized benchmark answer under `reference$extraction_target`; and
3. the selected LibeRary prediction (primary, text, vision, or reconciled).

```{r, eval=FALSE}
library_reference_shiny(
  corpus = "validation/liberary/aed-pkpd-reference/0.2.2",
  predictions = "validation/liberary/aed-pkpd-benchmark/text-current/predictions",
  source_dir = "AED_PKPD"
)

# The same application is available through the general launcher.
library_shiny(mode = "reference", reference_root = root,
              predictions = "benchmark/current/predictions",
              source_dir = "AED_PKPD")
```

The field table uses JSON pointers, so a decision remains unambiguous even
when labels repeat within THETA, OMEGA, covariance, or SIGMA arrays. Choosing a
LibeRary or custom value changes only the successor's normalized target. The
verbatim appendix row is immutable. Model and field decisions are exported as
separate audit tables before the successor corpus is created.

## Review without rewriting history

Curation produces a new corpus directory. A compact CSV can promote records
after source-paper review:

```text
reference_id,tier,training_eligible,review_status,notes
aed_phb_23018530_m01,B,TRUE,reviewed,Checked equation and parameter table
```

```{r, eval=FALSE}
library_reference_revise(
  root,
  "validation/liberary/aed-pkpd-reference/0.2.0",
  decisions = "review-decisions.csv",
  version = "0.2.0",
  curator = "reviewer-id"
)
```

Trying to make a test record training-eligible is an error. Test records can
still be promoted to A/B so they become stronger benchmark targets.

## Training exports

The extraction export pairs source-PDF text with the reviewed JSON answer. It
does not use the appendix as input. By default it exports only explicitly
training-eligible A/B records from train and validation partitions.

```{r, eval=FALSE}
library_reference_training_export(
  "validation/liberary/aed-pkpd-reference/0.2.0",
  source_dir = "AED_PKPD",
  output_dir = "training/aed-pkpd",
  tasks = c("extraction", "triage"),
  partitions = c("train", "validation")
)
```

Adjudication examples are not manufactured from the answer. They require real
stored text-lane and vision-lane candidates, source evidence, and a reviewed
resolution.

The optional packaged Python trainer validates every JSONL row before loading
PyTorch. Locate it from either a source or installed package with
`library_reference_training_files()`:

```text
python <trainer-path> \
  --train-file training/aed-pkpd/extraction_train.jsonl \
  --eval-file training/aed-pkpd/extraction_validation.jsonl \
  --base-model exact/hugging-face-base \
  --output-dir adapters/aed-extractor \
  --qlora
```

For an 8 GB GPU, focused article chunks and a compact base model are preferable
to full-paper, long-context training. A larger remote GPU can train the same
adapter export; the resulting adapter or merged GGUF can then be deployed
locally, provided it uses the exact matching base model.
