---
title: "LibeRary autonomous evidence workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{LibeRary autonomous evidence workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

LibeRary separates discovery, acquisition, document intelligence, independent
model extraction, reconciliation, catalogue publication, and scientific
qualification. This enables a largely self-running literature workflow without
misrepresenting machine agreement as human validation.

## Durable configuration

`library_home()` is outside the installed package, so reinstalls do not remove
catalogues, search manifests, triage backlogs, logs, PDFs, document bundles, or
model selections. Secrets stay in environment variables.

```{r, eval=FALSE}
library(LibeRary)
cfg <- ingest_load_config()
cfg$entrez$email <- "you@example.org"
library_save_config(cfg)
```

The ingestion GUI automatically saves these non-secret settings before each
job. The NCBI email identifies the E-utilities operator; it is not an API key.
An API key is optional at three requests/second or less. For higher rates, put
the key in `ENTREZ_KEY` (or enter it for the current GUI session); LibeRary does
not write secrets to its YAML configuration.

## Five selectable LLM roles

The roles are `triage`, `indexing` (investigation and synthesis), `vision`
(original PDF extraction and verification), `assessment` (skeptical evidence
review), and `adjudication`. Each can select an available
model from Ollama, OpenAI, or an OpenAI-compatible endpoint.

```{r, eval=FALSE}
library_llm_models("ollama", cfg, role = "indexing")

cfg$llm$triage$provider <- "same"
cfg$llm$indexing$model <- "text-model"
cfg$llm$vision$model <- "vision-model"
cfg$llm$adjudication$model <- "adjudication-model"
cfg$llm$require_independent_extraction_models <- FALSE
```

Every role may also define an `instruction` override. The ingestion GUI exposes
this beside the provider/model controls. Saving persists the override; restoring
the default removes it so the role follows the package instruction again.

Remote content transfer is disabled until
`cfg$llm$allow_remote_content <- TRUE`. This is an explicit data-transfer guard,
independent of the article-acquisition workflow.

## Search, triage, and backlog

```{r, eval=FALSE}
found <- ingest_discover(limit = 100, cfg = cfg)
found$summary
found$low_backlog_path
```

The triage LLM sees only the title and abstract. Its probability is converted
to deterministic thresholds. High and Intermediate records enter the default
first pass. Low records are not deleted or marked irrelevant: they are retained
both in the complete manifest and in a dedicated backlog CSV.

The backlog can later be acquired and processed explicitly:

```{r, eval=FALSE}
ingest_fetch_institutional(
  found$manifest_path, cfg = cfg,
  classes = "deferred_low", tiers = "low"
)
ingest_process_batch(found$manifest_path, cfg = cfg, tiers = "low")
```

## Canonical document bundles

`ingest_document_bundle()` stores an immutable source hash and creates a
content-addressed directory containing:

- the acquired source PDF;
- Docling standard-pipeline JSON, Markdown, and optional HTML;
- parser name, version, settings, logs, and fallback status;
- selected page images rendered directly from the original PDF; and
- the acquisition path and SHA-256 content identity.

Docling standard parsing is deliberately separate from the VLM lane. If
Docling is unavailable and fallback is enabled, LibeRary records a degraded
`pdftools_fallback` rather than pretending the layout-aware parse succeeded.

```{r, eval=FALSE}
ingest_docling_available(cfg)
bundle <- ingest_document_bundle(metadata, "article.pdf", cfg)
```

## Deliberative investigation, dual extraction, and reconciliation

The parsed-text path is not a one-shot extraction. It first maps all models
discussed in the article, then runs focused investigations of structure,
THETAs/covariates, OMEGA/IOV, SIGMA/observation model, population/dosing, and
reproduction data. A skeptical model challenges the evidence ledger, targeted
gap-search rounds revisit unresolved questions, and a vision model can verify
material claims against the original PDF pages. Deterministic completeness and
consistency checks run before an evidence-constrained synthesis produces the
standard LibeRary schema.

Each stage is content-addressed and cached beside the document bundle. The
ledger records reported, derived, inferred, conflicting, and unresolved claims
with source locators. A copy is preserved with every published catalogue
version. The legacy one-shot text path remains available by setting
`cfg$deliberative$enabled <- FALSE`.

The independent vision lane then reads selected original PDF pages, including
equations, tables, figures, captions, and footnotes. Both final lanes return the
same strict pharmacometric schema with evidence locators.

```{r, eval=FALSE}
dual <- ingest_dual_extract(metadata, bundle, cfg, adjudicate = TRUE)
dual$comparison$differences
dual$status
```

Claims such as model presence, canonical structure, ADVAN/TRANS, THETA, OMEGA,
SIGMA, covariates, population cohorts, demographics, CYP records, dosing, and
digitized concentration targets are compared field by field. A third configured
model adjudicates discrepancies against both source representations. Only
unresolved major fields require review by default.

The extraction schema stores both the publication's reported variability number
and its metric. The catalogue mapper converts CV and SD quantities to NONMEM
variance scale, distinguishes additive normal ETAs from exponential/log-normal
ETAs, preserves IIV versus IOV, and writes supplied correlations/covariances as
block OMEGA records. Ambiguous variability, shrinkage, RSEs, confidence
intervals, or bootstrap precision are not silently treated as OMEGA/SIGMA.

Terminal machine outcomes are:

- `machine_consistent`: both lanes agree on the material claims;
- `machine_adjudicated`: discrepancies were resolved without a remaining major
  conflict; and
- `needs_review`: a lane failed or a major field remains unresolved.

These are not equivalent to `validated`.

## Canonical implementation semantics

The extraction represents the mathematical structure before selecting a
NONMEM implementation: compartment count, input process, elimination type, and
parameterization are separate fields. Explicitly reported ADVAN/TRANS values
are retained as reported. In their absence, common linear one-, two-, and
three-compartment structures can be mapped to ADVAN1/2/3/4/11/12 with TRANS1,
TRANS2, or TRANS4 as appropriate. The mapping records provenance, confidence,
rationale, alternatives, and whether review remains required. Ambiguous or
nonlinear structures remain unresolved with general ODE implementations as
alternatives.

```{r, eval=FALSE}
enriched <- library_model_enrich(extraction)
enriched$structural_model$canonical
enriched$structural_model$implementations
```

Population data are cohort-aware. The original text is never discarded, and
mean/SD, median/quantiles, ranges, category counts, and individual values can
coexist for a descriptor. Dosing and pharmacogenetic evidence follow the same
source-locator and confidence convention.

## Conservative reproduction checks

Published catalogue entries contain a reproduction plan when the feature is
enabled. Auto-execution is off by default. The plan is deliberately blocked
when an implementation is unresolved, a dose or unit is missing, a normalized
dose lacks weight/BSA, a steady-state regimen lacks an interval, an infusion
lacks a duration, the model does not compile, or generated parameter defaults
would be required.

```{r, eval=FALSE}
plan <- library_reproduction_plan(enriched)
if (plan$eligible) {
  result <- library_reproduction_run(
    plan, "reproduction/pmid-123", nsim = 200, n_cores = 4
  )
}
```

When figure/table concentration targets are available, LibeRation simulates a
typical profile and a population distribution and writes the event datasets,
simulations, comparison metrics, audit JSON, and a persistent plot. A close
reproduction supports computational consistency under the recorded assumptions;
it is not independent scientific validation.

## End-to-end processing and publication

```{r, eval=FALSE}
result <- ingest_process_batch(
  found$manifest_path,
  cfg = cfg,
  tiers = c("high", "intermediate"),
  resume = TRUE,
  adjudicate = TRUE
)
library_list(root = cfg$catalog_dir)
```

Article decisions are saved beside the document bundle, so an interrupted run
can resume at its last valid investigation stage. Papers without a recoverable model receive an auditable exclusion
decision rather than an empty catalogue entry. Published models retain the
source hash, parser, dual-lane audit, comparison, and adjudication.

With `resume = TRUE`, only terminal decisions with the same source hash,
pipeline, and prompt version are reused, including review-required decisions.
Older one-shot decisions are therefore
upgraded automatically. Use `resume = FALSE` for a deliberate re-extraction, in
which case catalogue replacement uses the normal version-history path.

## LibeRation and LibeRties

Catalogue entries import through LibeRation's public control-stream parser and
compiler. The original catalogue identity and qualification state remain
attached to the model version.

```{r, eval=FALSE}
imported <- library_use_in_workspace(
  "pmid_12345678",
  workspace = LibeRation::nm_workspace()
)
```

The restricted LibeRties contract supports all pipeline stages. It cannot
contain functions, environments, external pointers, or API keys.

```{r, eval=FALSE}
job <- library_job(
  "dual_extract", metadata,
  pdf_path = "article.pdf",
  confirm_transfer = TRUE,
  cfg = cfg
)
queue$submit(job)
```

Workers return evidence results and never publish into a catalogue implicitly.
